---
title: "BF_comparison"
author: "Ulrich Lösener"
date: "9-9-2024"
output: html_document
---

In this file, I compare the results of the function `BF` from the package BFpack (v. 1.2.3), the function `bain` from the bain package (v. 0.2.8), and my own function (`bf_hand`) to calculate the multivariate Bayes Factor for informative hypotheses. 

First, I define a function to "translate" hypotheses into restriction matrices

```{r hyp2mat}

hyp2mat <- function(hypothesis) {
  
  # Split the hypothesis into parts
  parts <- unlist(strsplit(hypothesis, "(?=[<>=])|(?<=[<>=])", perl = TRUE))
  
  # Extract parameters
  parameters <- unique(parts[!parts %in% c("<", ">", "=")])
  num_params <- length(parameters)
  
  # storage for restriction matrices
  ineq_constraints <- list()
  eq_constraints <- list()
  
  # Parse the hypothesis
  for (i in seq(1, length(parts) - 1, by = 2)) {
    param1 <- parts[i]
    operator <- parts[i + 1]
    param2 <- parts[i + 2]
    
    index1 <- match(param1, parameters)
    index2 <- match(param2, parameters)
    
    if (operator == "<" || operator == ">") {
      constraint <- rep(0, num_params)
      if (operator == "<") {
        constraint[c(index1, index2)] <- c(-1, 1)
      } else {
        constraint[c(index1, index2)] <- c(1, -1)
      }
      ineq_constraints[[length(ineq_constraints) + 1]] <- constraint
    } else if (operator == "=") {
      constraint <- rep(0, num_params)
      constraint[c(index1, index2)] <- c(1, -1)
      eq_constraints[[length(eq_constraints) + 1]] <- constraint
    }
  }
  
  # Convert lists to matrices
  if (length(ineq_constraints) > 0) {
    Ineq.mat <- do.call(rbind, ineq_constraints)
  } else {
    Ineq.mat <- matrix(0, 0, num_params)
  }
  
  if (length(eq_constraints) > 0) {
    Eq.mat <- do.call(rbind, eq_constraints)
  } else {
    Eq.mat <- matrix(0, 0, num_params)
  }
  
  list(Ineq.mat = Ineq.mat, Eq.mat = Eq.mat)
}
```

Next, I define my own function to calculate BF_u for informative hypotheses. 

```{r my own multivariate BF function}

bf_hand <- function(N, est, sigma, hypothesis, fraction = 1) {
  
  hypothesis <- gsub("([>=<])", " \\1 ", hypothesis) # Add spaces
  hypothesis <- gsub("\\s+", " ", hypothesis) # Remove any extra spaces
  hypothesis <- trimws(hypothesis) # Trim leading and trailing spaces

  b <- fraction/N
  struc <- unlist(strsplit(hypothesis, " "))
  params <- struc[seq(1, length(struc), 2)]
  constr <- struc[seq(2, length(struc) - 1, 2)]
  ineq.constr <- constr[constr == "<" | constr == ">"]
  eq.constr <- constr[constr == "="]
  
  # Contrast matrix
  Rmat <- hyp2mat(hypothesis = hypothesis)
  Rmat_ineq <- Rmat$Ineq.mat
  Rmat_eq <- Rmat$Eq.mat
  
  # Compute the complexity (c) and fit (f) for inequalities
  if(!is.null(Rmat_ineq) && nrow(Rmat_ineq) > 0) {
    
    if(any(constr == "<")) {
      est_adjusted <- -est
    } else {
      est_adjusted <- est
    }
    
    # Initialize bounds for inequalities
    lower <- rep(-Inf, length(ineq.constr))
    upper <- rep(Inf, length(ineq.constr))
    
    # Assign bounds based on inequality constraints
    lower[ineq.constr == ">"] <- 0
    upper[ineq.constr == "<"] <- 0
    
    c_ineq <- as.numeric(mvtnorm::pmvnorm(lower = lower, upper = upper,
                                          mean = rep(0, length(ineq.constr)), 
                                          sigma = Rmat_ineq %*% sigma %*% t(Rmat_ineq)/b,
                                          keepAttr = F))
    f_ineq <- as.numeric(mvtnorm::pmvnorm(lower = lower, upper = upper,
                                          mean = c(Rmat_ineq %*% est_adjusted),
                                          sigma = Rmat_ineq %*% sigma %*% t(Rmat_ineq),
                                          keepAttr = F))
    BFu_ineq <- f_ineq / c_ineq
  } else {
    c_ineq <- 1
    f_ineq <- 1
    BFu_ineq <- 1
  }
  
  if (!is.null(Rmat_eq) && nrow(Rmat_eq) > 0) {
    c_eq <- mvtnorm::dmvnorm(rep(0, nrow(Rmat_eq)), 
                             sigma = Rmat_eq %*% sigma %*% t(Rmat_eq)/b)
    f_eq <- mvtnorm::dmvnorm(x = rep(0, nrow(Rmat_eq)), mean = Rmat_eq %*% est, 
                             sigma = Rmat_eq %*% sigma %*% t(Rmat_eq))
    BFu_eq <- f_eq / c_eq
  } else {
    c_eq <- 1
    f_eq <- 1
    BFu_eq <- 1
  }
  
  BFu <- BFu_ineq * BFu_eq
  
  return(list(complex_ineq = c_ineq, 
              complex_eq = c_eq,
              fit_ineq = f_ineq, 
              fit_eq = f_eq,
              BFu = BFu))
}
```

Now, we load the packages bain and BFpack.

```{r}
library(bain)
library(BFpack)
```

### Simulation 1: Fix N, vary effect size

The effect size in this case is the difference between the three parameters "a", "b", and "c". In this simulation, we evaluate 100 different effect sizes from very small (a=0, b=0.1, c=0.2) to very large (a=0, b=10, c=20), keeping N constant at N=100.

The first hypothesis under consideration is $H_1: a<b<c$. We expect the $BF_{1u}$ to increase with increasing effect size. 

```{r simulation1}

hyp1 <- "a<b<c" # define our hypothesis H1

sig <- matrix(c(1, 0.3, 0.3, # covariance matrix of parameters
                0.3, 1, 0.3,
                0.3, 0.3, 1), nrow = 3)


m <- 100 # number of iterations
ests <- list() # space for estimates of a, b, and c

# create vectors of a, b, and c according to different effect sizes and store them in ests
for(i in 1:m){ 
  a <- 0
  b <- a+i/10
  c <- b+i/10
  ests[[i]] <- c(a, b, c)
  names(ests[[i]]) <- c("a", "b", "c")
}

# storage
BFs_hand <- rep(NA, m)
BFs_bfpack <- rep(NA, m)
BFs_bain <- rep(NA, m)

# fix N, vary eff.size
set.seed(123)
for(i in 1:m){
  r <- bf_hand(est = ests[[i]], N=100, sigma=sig, hypothesis=hyp1)
  BFs_hand[i] <- r$BFu
  
  r2 <- BF(x=ests[[i]], n=100, Sigma=sig, hypothesis=hyp1)
  BFs_bfpack[i] <- r2[["BFtable_confirmatory"]][1,7]
  
  r3 <- bain(x=ests[[i]], n=100, Sigma=sig, hypothesis=hyp1)
  BFs_bain[i] <- r3[["fit"]][["BF.u"]][1]
}

```

Now, we plot the results.

```{r plots1}
library(ggplot2)

sim_dat1 <- data.frame(cbind(c(BFs_hand, BFs_bfpack, BFs_bain), rep(c("hand", "bfpack", "bain"), each=100), as.numeric(rep(seq(1:100), 3))))
names(sim_dat1) <- c("bf", "type", "iteration") 
sim_dat1$bf <- as.numeric(sim_dat1$bf)
sim_dat1$iteration <- as.numeric(sim_dat1$iteration)

ggplot(sim_dat1, aes( x=iteration, y=bf, color=type)) +
  geom_line() +
  xlab("effect size") +
  ggtitle("H1: a<b<c")

BFs_bfpack == BFs_hand # TRUE
```

My function `bf_hand` gives the exact same results as BFpack. The results from bain are a bit noisy but also in accordance with the other two functions.

Now, we do the same for a different hypothesis. $H_2:a<b=c$. Because we know that $b\neq c$, we expect the BF to decline at some point. For convenience, the code for the second simulation is hidden. It is very similar to the code for the first simulation with exception of the hypothesis.

```{r simulation1b, include=FALSE}

hyp2 <- "a<b=c" # define our hypothesis H1

sig <- matrix(c(1, 0.3, 0.3, # covariance matrix of parameters
                0.3, 1, 0.3,
                0.3, 0.3, 1), nrow = 3)


m <- 100 # number of iterations
ests <- list() # space for estimates of a, b, and c

# create vectors of a, b, and c according to different effect sizes and store them in ests
for(i in 1:m){ 
  a <- 0
  b <- a+i/10
  c <- b+i/10
  ests[[i]] <- c(a, b, c)
  names(ests[[i]]) <- c("a", "b", "c")
}

# storage
BFs_hand <- rep(NA, m)
BFs_bfpack <- rep(NA, m)
BFs_bain <- rep(NA, m)

# fix N, vary eff.size
set.seed(123)
for(i in 1:m){
  r <- bf_hand(est = ests[[i]], N=100, sigma=sig, hypothesis=hyp2)
  BFs_hand[i] <- r$BFu
  
  r2 <- BF(x=ests[[i]], n=100, Sigma=sig, hypothesis=hyp2)
  BFs_bfpack[i] <- r2[["BFtable_confirmatory"]][1,7]
  
  r3 <- bain(x=ests[[i]], n=100, Sigma=sig, hypothesis=hyp2)
  BFs_bain[i] <- r3[["fit"]][["BF.u"]][1]
}

```

Now, we plot the results.

```{r plots2}
sim_dat1 <- data.frame(cbind(c(BFs_hand, BFs_bfpack, BFs_bain), rep(c("hand", "bfpack", "bain"), each=100), as.numeric(rep(seq(1:100), 3))))
names(sim_dat1) <- c("bf", "type", "iteration") 
sim_dat1$bf <- as.numeric(sim_dat1$bf)
sim_dat1$iteration <- as.numeric(sim_dat1$iteration)

ggplot(sim_dat1, aes( x=iteration, y=bf, color=type)) +
  geom_line() +
  xlab("effect size") +
  ggtitle("H2: a<b=c")

```

This is interesting. All three functions give different results. My own function `bf_hand` seems to be in between the other two. 
Let's see what happens when the hypothesis is $H_3:a=b=c$

```{r simulation1c, include=FALSE}

hyp3 <- "a=b=c" # define our hypothesis H1

sig <- matrix(c(1, 0.3, 0.3, # covariance matrix of parameters
                0.3, 1, 0.3,
                0.3, 0.3, 1), nrow = 3)


m <- 100 # number of iterations
ests <- list() # space for estimates of a, b, and c

# create vectors of a, b, and c according to different effect sizes and store them in ests
for(i in 1:m){ 
  a <- 0
  b <- a+i/10
  c <- b+i/10
  ests[[i]] <- c(a, b, c)
  names(ests[[i]]) <- c("a", "b", "c")
}

# storage
BFs_hand <- rep(NA, m)
BFs_bfpack <- rep(NA, m)
BFs_bain <- rep(NA, m)

# fix N, vary eff.size
set.seed(123)
for(i in 1:m){
  r <- bf_hand(est = ests[[i]], N=100, sigma=sig, hypothesis=hyp3)
  BFs_hand[i] <- r$BFu
  
  r2 <- BF(x=ests[[i]], n=100, Sigma=sig, hypothesis=hyp3)
  BFs_bfpack[i] <- r2[["BFtable_confirmatory"]][1,7]
  
  r3 <- bain(x=ests[[i]], n=100, Sigma=sig, hypothesis=hyp3)
  BFs_bain[i] <- r3[["fit"]][["BF.u"]][1]
}

```


```{r plots3, echo=-2}
sim_dat1 <- data.frame(cbind(c(BFs_hand, BFs_bfpack, BFs_bain), rep(c("hand", "bfpack", "bain"), each=100), as.numeric(rep(seq(1:100), 3))))
names(sim_dat1) <- c("bf", "type", "iteration") 
sim_dat1$bf <- as.numeric(sim_dat1$bf)
sim_dat1$iteration <- as.numeric(sim_dat1$iteration)

ggplot(sim_dat1, aes(x=iteration, y=bf, color=type)) +
  geom_line() +
  xlab("effect size") +
  ggtitle("H3: a=b=c")

BFs_bfpack == BFs_hand # TRUE
```

As was the case in the first simulation, the function from BFpack and `bf_hand` give the exact same results. Interestingly, the BF from bain is exactly half of the other BFs. 
